Refinement of Thermostated Molecular Dynamics Using Backward Error Analysis

### 2019-02-25 rigid_bodies.tex
include braun reference

Molecular dynamics (MD) simulation involves several types of approximations in the quest for mimicking real systems.
Naively, we tend to rank simulation results by their proximity to experimental data and attribute any disagreement to a lack of accuracy of the employed force field.
However, systematic errors can possibly be hidden in these deviations, being the cause of non-physical behavior.
It has been recently shown \cite{Palmer_2018} that erroneous simulation results \cite{Limmer_2011, Limmer_2013} were the cause of a long-standing controversy \cite{Smart_2018} about a liquid-liquid transition believed to exist for supercooled water.
In this case, hybrid Monte Carlo steps were employed with initial velocities that failed to comply with the equipartition between translational and rotational degrees of freedom \cite{Palmer_2018}.
As a matter of fact, similar ill-partitioning of kinetic energy can arise in MD simulations as an artifact due to the discretization of the equations of motion.
This can affect not only the translation/rotation equipartition in rigid-body systems \cite{Davidchack_2010, Silveira_2017}, but also the distribution of energy among molecules of distinct sizes \cite{Eastwood_2010}.
The issue with rigid-body MD is relevant not only because water \cite{Jorgensen_1983} and other small molecules are usually modeled as rigid bodies, but also because designing proteins, nanoparticles, or other large structures as collections of interconnected bodies \cite{Miller_2002, Knorowski_2012, Patra_2013} can be seen as a useful coarse-graining strategy.
It also brings into question the NVT-MD method developed by Kamberaj, Low, and Neal \cite{Kamberaj_2005} (KLN) who, by assuming equipartition, coupled two independent thermostats to the translational and rotational degrees of freedom.
As we will show here, the method fails to reproduce the specified temperature even for time-step sizes typically employed in MD.
In fact, Davidchack \cite{Davidchack_2010} recommends that the use of separate thermostats for translation and rotation should always be avoided.
Nonetheless, the cited double-thermostat method is available in software packages such as HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995} and has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, membranes \cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles \cite{Yan_2008}, and nanoparticles \cite{Patra_2014}, in spite of all the issues observed here.

>>>>

Molecular dynamics (MD) simulation involves several types of approximations in the quest for mimicking real systems.
Naively, we tend to rank simulation results by their proximity to experimental data and attribute any disagreement to a lack of accuracy of the employed force field.
However, systematic errors can possibly be hidden in these deviations, being the cause of non-physical behavior.
It has been recently shown \cite{Palmer_2018} that erroneous simulation results \cite{Limmer_2011, Limmer_2013} were the cause of a long-standing controversy \cite{Smart_2018} about a liquid-liquid transition believed to exist for supercooled water.
In this case, hybrid Monte Carlo steps were employed with initial velocities that failed to comply with the equipartition between translational and rotational degrees of freedom \cite{Palmer_2018}.
As a matter of fact, similar ill-partitioning of kinetic energy can arise in MD simulations coupled with noncanonical schemes \cite{Braun_2018}, or as an artifact due to the discretization of the equations of motion.
This can affect not only the translation/rotation equipartition in rigid-body systems \cite{Davidchack_2010, Silveira_2017}, but also the distribution of energy among molecules of distinct sizes \cite{Eastwood_2010}.
The issue with rigid-body MD is relevant not only because water \cite{Jorgensen_1983} and other small molecules are usually modeled as rigid bodies, but also because designing proteins, nanoparticles, or other large structures as collections of interconnected bodies \cite{Miller_2002, Knorowski_2012, Patra_2013} can be seen as a useful coarse-graining strategy.
It also brings into question the NVT-MD method developed by Kamberaj, Low, and Neal \cite{Kamberaj_2005} (KLN) who, by assuming equipartition, coupled two independent thermostats to the translational and rotational degrees of freedom.
As we will show here, the method fails to reproduce the specified temperature even for time-step sizes typically employed in MD.
In fact, Davidchack \cite{Davidchack_2010} recommends that the use of separate thermostats for translation and rotation should always be avoided.
Nonetheless, the cited double-thermostat method is available in software packages such as HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995} and has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, membranes \cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles \cite{Yan_2008}, and nanoparticles \cite{Patra_2014}, in spite of all the issues observed here.
